---
title: "Cell type classification in single cell datasets"
author: "Mike Tyler"
date: "07/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load_packages_and_data}
# To render this R markdown file:
# rmarkdown::render('sc_data_cell_type_classification.rmd', output_dir = '../data_and_figures/sc_reclassify')

library(data.table) # 1.12.8
library(ggplot2) # This is version 3.3.1 on my laptop's WSL setup but 3.3.0 on WEXAC.
library(magrittr) # 1.5
library(plyr) # 1.8.6
library(cowplot) # 1.0.0

source('general_functions.R')
```





# {.tabset}





## Breast - Qian

```{r}
plot_data <- fread('../data_and_figures/sc_reclassify/files_for_rmd/breast_qian_plot_data.csv')
de_11_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/breast_qian_de_11_top_50.rds')
de_caf_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/breast_qian_de_caf_top_50.rds')
de_between_caf_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/breast_qian_de_between_caf_top_50.rds')

# Scatter plot of t-SNE coordinates, coloured by DBSCAN cluster:
tsne_plot_dbscan <- ggplot(plot_data, aes(x = x, y = y, colour = dbscan_cluster)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$dbscan_cluster)))) +
	theme_minimal() +
	labs(title = 'Clustering by t-SNE and DBSCAN')

tsne_plot_author_cell_types <- ggplot(plot_data, aes(x = x, y = y, colour = cell_type)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$cell_type)))) +
	theme_minimal() +
	labs(title = "Authors' cell type classifications")
```

```{r fig.width=12, fig.height=5}
plot_grid(tsne_plot_dbscan, tsne_plot_author_cell_types, nrow = 1, ncol = 2)
```

The DBSCAN clusters agree well with the cell type classifications from the authors, with the exception of cluster 11.
The authors identify these as dendritic cells, but I do not detect high expression of DC markers in this cluster.
Also, the CAFs appear to divide into two main DBSCAN clusters, 2 and 5.
So we want to identify cluster 11 and check the difference between clusters 2 and 5.

The top 50 differentially expressed genes in cluster 11 contain various immune markers:

```{r}
de_11_top_50
```

Here are the top 50 differentially expressed genes between each of clusters 2 and 5 and all the other clusters:

```{r}
de_caf_top_50
```

Here are top 50 differentially expressed genes between clusters 2 and 5 (ignoring all other clusters):

```{r}
de_between_caf_top_50
```

It looks like cluster 5 could be pericytes, since RGS5 comes up prominently in this cluster.





## CRC - Lee (SMC)

```{r}
plot_data <- fread('../data_and_figures/sc_reclassify/files_for_rmd/crc_lee_smc_plot_data.csv')
de_5_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/crc_lee_smc_de_5_top_50.rds')
de_caf_endothelial_top_50<- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/crc_lee_smc_de_caf_endothelial_top_50.rds')
de_between_caf_endothelial_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/crc_lee_smc_de_between_caf_endothelial_top_50.rds')
de_noise_2_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/crc_lee_smc_de_noise_2_top_50.rds')
de_noise_2_caf_endothelial_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/crc_lee_smc_de_noise_2_caf_endothelial_top_50.rds')

# Scatter plot of t-SNE coordinates, coloured by DBSCAN cluster:
tsne_plot_dbscan <- ggplot(plot_data, aes(x = x, y = y, colour = dbscan_cluster)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$dbscan_cluster)))) +
	theme_minimal() +
	labs(title = 'Clustering by t-SNE and DBSCAN')

tsne_plot_author_cell_types <- ggplot(plot_data, aes(x = x, y = y, colour = cell_type)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$cell_type)))) +
	theme_minimal() +
	labs(title = "Authors' cell type classifications")
```

```{r fig.width=12, fig.height=5}
plot_grid(tsne_plot_dbscan, tsne_plot_author_cell_types, nrow = 1, ncol = 2)
```

The DBSCAN clusters broadly agree with the cell type classifications from the authors, with a couple of exceptions.
Firstly, my analysis identifies cluster 2 as CAFs and cluster 3 as endothelial cells, but these are both classified as fibroblasts by the authors.
Secondly, cluster 5 appears very distinct from cluster 4 (the macrophage cluster), but the two are grouped by the authors into a single "myeloid" cluster.
Thirdly, there are a couple of small clusters labelled as "noise" by the DBSCAN algorithm, but which are classified respectively as B cells and fibroblasts by the authors.
So we want to investigate these clusters to see whether we agree with the classifications from the authors.

The top 50 differentially expressed genes in cluster 5 contain various immune markers:

```{r}
de_5_top_50
```

Here are the top 50 differentially expressed genes between each of clusters 2 and 3 and all the other clusters:

```{r}
de_caf_endothelial_top_50
```

Here are top 50 differentially expressed genes between clusters 2 and 3 (ignoring all other clusters):

```{r}
de_between_caf_endothelial_top_50
```

This indeed suggests that cluster 3 does not consist of CAFs but more likely endothelial cells or pericytes.

We also test the group of "noise" cells that cluster close to the candidate endothelial cells.
Here are the top 50 differentially expressed genes between this group of "noise" cells and all other clusters:

```{r}
de_noise_2_top_50
```

Here are the top 50 differentially expressed genes between these "noise" cells and the candidate CAF and endothelial clusters (ignoring all other clusters):

```{r}
de_noise_2_caf_endothelial_top_50
```

We see RGS5 and PECAM1, suggesting that these "noise" cells may be pericytes.

A similar test for the other group of "noise" cells confirms that they are indeed B cells, as classified by the authors (not shown).





## Liver - Ma

```{r}
plot_data <- fread('../data_and_figures/sc_reclassify/files_for_rmd/liver_ma_plot_data.csv')
de_caf_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/liver_ma_de_caf_top_50.rds')
de_endothelial_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/liver_ma_de_endothelial_top_50.rds')
de_between_endothelial_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/liver_ma_de_between_endothelial_top_50.rds')

# Scatter plot of t-SNE coordinates, coloured by DBSCAN cluster:
tsne_plot_dbscan <- ggplot(plot_data, aes(x = x, y = y, colour = dbscan_cluster)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$dbscan_cluster)))) +
	theme_minimal() +
	labs(title = 'Clustering by t-SNE and DBSCAN')

tsne_plot_author_cell_types <- ggplot(plot_data, aes(x = x, y = y, colour = cell_type)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$cell_type)))) +
	theme_minimal() +
	labs(title = "Authors' cell type classifications")
```

```{r fig.width=12, fig.height=5}
plot_grid(tsne_plot_dbscan, tsne_plot_author_cell_types, nrow = 1, ncol = 2)
```

The DBSCAN clusters do not effectively separate the cell types, although the marker gene expression broadly agrees with the cell type classifications from the authors.
However, the authors classified a large number of cells as endothelial, including DBSCAN clusters 5 and 6 and some of 2.
We will test these groups, along with the remainder of cluster 2, to check whether they correspond to CAFs or endothelial cells, or something similar.

Here are the top 50 differentially expressed genes in the candidate CAF cluster (cells in cluster 2 not classified as endothelial by the authors):

```{r}
de_caf_top_50
```

These cells look rather like myofibroblasts, or even myocytes.

Here are the top 50 differentially expressed genes between each of the three candidate endothelial cell groups (clusters 5 and 6 and some of 2) and all the other clusters:

```{r}
de_endothelial_top_50
```

Here are top 50 differentially expressed genes between these three groups (ignoring all other clusters):

```{r}
de_between_endothelial_top_50
```

This is hard to interpret, except that cluster 6 looks more myocyte-like.
This cluster also has relatively high complexity, though this may not be relevant.





## Lung - Qian

```{r}
plot_data <- fread('../data_and_figures/sc_reclassify/files_for_rmd/lung_qian_plot_data.csv')
de_8_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/lung_qian_de_8_top_50.rds')
de_caf_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/lung_qian_de_caf_top_50.rds')
de_between_caf_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/lung_qian_de_between_caf_top_50.rds')

# Scatter plot of t-SNE coordinates, coloured by DBSCAN cluster:
tsne_plot_dbscan <- ggplot(plot_data, aes(x = x, y = y, colour = dbscan_cluster)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$dbscan_cluster)))) +
	theme_minimal() +
	labs(title = 'Clustering by t-SNE and DBSCAN')

tsne_plot_author_cell_types <- ggplot(plot_data, aes(x = x, y = y, colour = cell_type)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$cell_type)))) +
	theme_minimal() +
	labs(title = "Authors' cell type classifications")
```

```{r fig.width=12, fig.height=5}
plot_grid(tsne_plot_dbscan, tsne_plot_author_cell_types, nrow = 1, ncol = 2)
```

The DBSCAN clusters agree well with the cell type classifications from the authors, with some minor exceptions.
Firstly, cluster 8 is part of the cluster identified as "myeloid" by the authors, but I do not detect high expression of any of my cell type markers in this cluster.
Secondly, some of the cells classified as "myeloid" by the authors appear in the B cell clusters (DBSCAN clusters 4, 13, 18, and 19)
Also, the CAFs appear to divide into three main DBSCAN clusters, 3, 14 and 15.
So we want to identify cluster 8 and check the difference between clusters 3, 14 and 15 (and possibly reclassify some "myeloid" cells as B cells).

The top 50 differentially expressed genes in cluster 8 contain various immune markers:

```{r}
de_8_top_50
```

Here are the top 50 differentially expressed genes between each of clusters 3, 14 and 15 and all the other clusters:

```{r}
de_caf_top_50
```

Here are top 50 differentially expressed genes between clusters 3, 14 and 15 (ignoring all other clusters):

```{r}
de_between_caf_top_50
```

The clusters all look CAF-like, though 14 and 15 look more myocyte-like, and RGS5 and PDGFRB appear in cluster 15.
Thus, perhaps cluster 14 could be classified as myofibroblasts, while 15 could consist of pericytes.





## Ovarian - Qian

```{r}
plot_data <- fread('../data_and_figures/sc_reclassify/files_for_rmd/ovarian_qian_plot_data.csv')
de_9_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/ovarian_qian_de_9_top_50.rds')
de_caf_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/ovarian_qian_de_caf_top_50.rds')
de_between_caf_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/ovarian_qian_de_between_caf_top_50.rds')
de_endothelial_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/ovarian_qian_de_endothelial_top_50.rds')
de_between_endothelial_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/ovarian_qian_de_between_endothelial_top_50.rds')

# Scatter plot of t-SNE coordinates, coloured by DBSCAN cluster:
tsne_plot_dbscan <- ggplot(plot_data, aes(x = x, y = y, colour = dbscan_cluster)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$dbscan_cluster)))) +
	theme_minimal() +
	labs(title = 'Clustering by t-SNE and DBSCAN')

tsne_plot_author_cell_types <- ggplot(plot_data, aes(x = x, y = y, colour = cell_type)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$cell_type)))) +
	theme_minimal() +
	labs(title = "Authors' cell type classifications")
```

```{r fig.width=12, fig.height=5}
plot_grid(tsne_plot_dbscan, tsne_plot_author_cell_types, nrow = 1, ncol = 2)
```

The DBSCAN clusters agree well with the cell type assignments from the authors, though some cell types appear separated into very distinct clusters.
Firstly, DBSCAN clusters 9 and 13 are very distinct, and I do not detect a strong B cell signal in 9, but both are classified as B cells by the authors.
Secondly, the endothelial cells appear to separate into two distinct clusters - DBSCAN cluster 6 plus another group comprising DBSCAN clusters 5 and 14.
Thirdly, there appear to be many CAFs, but they are separated into several clusters.
So we want to identify cluster 9 and check for differences between the endothelial and CAF clusters, respectively.
We consider DBSCAN clusters 1, 2, 4 and 10 as a single CAF cluster, since there is no clear separation between them.

The top 50 differentially expressed genes in cluster 9 contain various immune markers:

```{r}
de_9_top_50
```

Here are the top 50 differentially expressed genes between each of the CAF clusters and all the other clusters:

```{r}
de_caf_top_50
```

Here are top 50 differentially expressed genes between the CAF clusters (ignoring all other clusters):

```{r}
de_between_caf_top_50
```

From this, I would guess cluster 8 consists of pericytes or similar.
Clusters 1, 2, 4, 10, 18 and 20 all look like CAFs, with some perhaps a little more myocyte-like than others.
Clusters 15 and 17 are harder to identify.
In cluster 17 I see GZMA and GZMK, suggesting an immune element.
In cluster 15 I see a lot of ribosomal proteins - this cluster has somewhat low complexity, but not drastically.

Here are the top 50 differentially expressed genes between each of the endothelial clusters and all the other clusters:

```{r}
de_endothelial_top_50
```

Here are top 50 differentially expressed genes between the endothelial clusters (ignoring all other clusters):

```{r}
de_between_endothelial_top_50
```

I cannot see any meaningful differences between the endothelial clusters, so I think it is fine to classify them all as endothelial cells.





## PDAC - Peng

```{r}
plot_data <- fread('../data_and_figures/sc_reclassify/files_for_rmd/pdac_peng_plot_data.csv')
de_caf_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/pdac_peng_de_caf_top_50.rds')
de_between_caf_top_50 <- readRDS('../data_and_figures/sc_reclassify/files_for_rmd/pdac_peng_de_between_caf_top_50.rds')

# Scatter plot of t-SNE coordinates, coloured by DBSCAN cluster:
tsne_plot_dbscan <- ggplot(plot_data, aes(x = x, y = y, colour = dbscan_cluster)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$dbscan_cluster)))) +
	theme_minimal() +
	labs(title = 'Clustering by t-SNE and DBSCAN')

tsne_plot_author_cell_types <- ggplot(plot_data, aes(x = x, y = y, colour = cell_type)) +
	geom_point() +
	scale_colour_manual(values = randomcoloR::distinctColorPalette(length(unique(plot_data$cell_type)))) +
	theme_minimal() +
	labs(title = "Authors' cell type classifications")
```

```{r fig.width=12, fig.height=5}
plot_grid(tsne_plot_dbscan, tsne_plot_author_cell_types, nrow = 1, ncol = 2)
```

The clustering by DBSCAN agrees well with the cell type classifications from the authors, except DBSCAN cluster 10 is grouped with the T cells by the authors but looks like B plasma cells in my analysis.
Also, since the CAF/stellate cells (DBSCAN clusters 4, 7, 12, 13 and 14) appear to separate into a few distinct clusters, we would like to check the differences between them.
We also checked that cluster 8 indeed consists of "endocrine" cells (not shown).

Here are the top 50 differentially expressed genes between each of the CAF clusters and all the other clusters:

```{r}
de_caf_top_50
```

Here are top 50 differentially expressed genes between the CAF clusters (ignoring all other clusters):

```{r}
de_between_caf_top_50
```

Clusters 4 and 12 indeed look like stellate cells, while 7, 13 and 14 indeed look like CAFs.
Cluster 14 looks a little different from 7 and 13, as expected (since it is very separate in the t-SNE), but it still strongly resembles CAFs.
In summary, it seems safe to use the classifications from the authors after possibly redefining cluster 10 as B plasma cells.
